Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-07-18 Compare_053014_statsmed_revisions.tex
Submission time

We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with an integrated $STG$, $\widetilde{STG}$, of 0.37, two marginal quality surrogates with $\widetilde{STG}$ of 0.1 and 0.25, and a null surrogate with $\widetilde{STG} = 0$. We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $\widetilde{STG}$ of 0.46 and 0.81, respectively. We investigate these different levels of candidate surrogates for two different BIP correlations; $R_{s_{1j}w}$, of 0.8 and 0.5 between the univariate candidate and univariate BIP. For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0,0,0.8), (0.5,0,0,0.5)\}$ between the bivariate candidate PS and the bivariate BIP $w=(w_1, w_1)$. We investigate correlations between the biomarkers in the bivariate PS ranging from 0 to 0.3 and the correlation between the Ws in the vector BIP from 0 to 1. Results for $R_{S_kS_j}=0.3$ and two independent Ws for a bivariate BIP are displayed in Tables \ref{Bias} and \ref{IPow}; other scenarios can be found in Appendix B of the supplementary materials. 500 replicates of the simulation were run for each scenario. All candidate PSs in the simulations are time invariant; results for time-dependent models can be found in Table 6, Appendix B of the supplementary materials.

We fit the Weibull risk in each scenario using the proposed semi-parametric EML method, censoring the imputations when the data generation was censored and using the imputations as generated if normal generation was used; scenarios indicated by a $\#$ in Tables \ref{Bias} and \ref{IPow} use censored generation and imputation. Bias results are displayed in Table \ref{Bias} and suggest that the method is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable. As we have suggested above, there is some bias when there is no surrogate value. When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed). A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves. Although unbiased, the results from using a univariate BIP for a multivariate candidate PS indicate that convergence issues can occur in some scenarios and that vector BIPs improve performance; results can be seen in Table 1 of Appendix B of the supplementary materials.

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
\begin{tabular}{lccccccc}
  \hline
&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG^{\#}(3)=0.48$&$STG(3)^{\#*}=0.23$&$STG(3)^{\#*}=0.51$\\ 
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
 &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
 &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
 &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & 0.023 (0.093) & 0.032 (0.054) \\
 & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.022 (0.095) & 0.069 (0.092) \\
Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   \hline
\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$. *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,0,0,x)$ for bivariate candidate PS and the bivariate BIP. \# indicates that the candidate surrogate was generated using a censored normal distribution.}
\end{tabular}
\end{center}
\end{sidewaystable}

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power to detect difference in $\widetilde{STG}$ \label{IPow}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}^{\#}=0.47$& $\widetilde{STG}^{\#*}=0.23$  & $\widetilde{STG}^{\#*}=0.52$\\
  \hline
$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 99.6 & 7.8 & 99.2 \\ 
&  0.5  & 5.7 & 17.8 & 99.8 & 9.2 & 99.4 \\ 
 $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 91.6 & 6.6 & 83.6 \\ 
&  0.5 &  & 5.5 & 95.2 & 7.1 & 89.8 \\ 
$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 82.6 & 8.8 \\ 
 & 0.5 &  &  & 5.2 & 81.9 & 16.2 \\ 
$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 75.8 \\ 
&  (0.5, 0.5) &  &  &  & 6.3 & 78.0 \\ 
$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.6 \\ 
 & (0.5, 0.5) &  &  &  &  & 5.9 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$. \# indicates that the candidate surrogate was generated using a censored normal distribution.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001. The P-value for testing H0$_1$ for EGFR is 0.18. The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.062. The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of $<$ 0.001. Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone. The estimated $\widehat{\widetilde{STG}}$ is 0.37 for HBA1C alone and 0.12 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.46. Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.

>>>>

We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with an integrated $STG$, $\widetilde{STG}$, of 0.47, two marginal quality surrogates with $\widetilde{STG}$ of 0.1 and 0.25, and a null surrogate with $\widetilde{STG} = 0$. We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $\widetilde{STG}$ of 0.23 and 0.52, respectively. We investigate these different levels of candidate surrogates for two different BIP correlations; $R_{s_{1j}w}$, of 0.8 and 0.5 between the univariate candidate and univariate BIP. For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0,0,0.8), (0.5,0,0,0.5)\}$ between the bivariate candidate PS and the bivariate BIP $w=(w_1, w_1)$. We investigate correlations between the biomarkers in the bivariate PS ranging from 0 to 0.3 and the correlation between the Ws in the vector BIP from 0 to 1. Results for $R_{S_kS_j}=0.3$ and two independent Ws for a bivariate BIP are displayed in Tables \ref{Bias} and \ref{IPow}; other scenarios can be found in Appendix B of the supplementary materials. 500 replicates of the simulation were run for each scenario. All candidate PSs in the simulations are time invariant; results for time-dependent models can be found in Table 6, Appendix B of the supplementary materials.

We fit the Weibull risk in each scenario using the proposed semi-parametric EML method, censoring the imputations when the data generation was censored and using the imputations as generated if normal generation was used; scenarios indicated by a $\#$ in Tables \ref{Bias} and \ref{IPow} use censored generation and imputation. Bias results are displayed in Table \ref{Bias} and suggest that the method is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable. As we have suggested above, there is some bias when there is no surrogate value. When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed). For two of the univariate scenarios the models are estimated under minor violations of assumption A6 (indicated by $\dagger$). In this case, the TPF and FPF correspond to the modified versions defined in section \ref{CAM}. A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves. Although unbiased, the results from using a univariate BIP for a multivariate candidate PS indicate that convergence issues can occur in some scenarios and that vector BIPs improve performance; results can be seen in Table 1 of Appendix B of the supplementary materials.

@@

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
\begin{tabular}{lccccccc}
  \hline
&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG^{\#}(3)=0.48$&$STG(3)^{\#*}=0.23$&$STG(3)^{\#*}=0.51$\\ 
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0.003 (0.029) & 0.002 (0.012) \\ 
 &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & 0.001 (0.024) & 0.003 (0.014) \\ 
Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058)$^\dagger$ & -0.024 (0.065)$^\dagger$ & 0.013 (0.048) & 0.005 (0.060) & 0.003 (0.019) \\ 
 &0.5& 0.56 (0.048) & 0.028 (0.07)$^\dagger$ & -0.006 (0.075)$^\dagger$ & 0.011 (0.047) & 0.029 (0.062) & 0.009 (0.022) \\ 
Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002)$^\dagger$ & 0.001 (0.003)$^\dagger$ & 0 (0.001) & 0.001 (0.005) & -0.006 (0.019) \\ 
 &0.5& 0.497 (0.002) & -0.001 (0.003)$^\dagger$ & 0 (0.003)$^\dagger$ & -0.001 (0.001) & -0.001 (0.024) & -0.004 (0.018) \\ 
Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & -0.005 (0.094) & 0.013 (0.026) \\
  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.028 (0.088) & 0.046 (0.034) \\ 
Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & -0.004 (0.096) & 0.010 (0.031) \\
 & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.032 (0.091) & 0.021 (0.04) \\
Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   \hline
\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$. *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,0,0,x)$ for bivariate candidate PS and the bivariate BIP. \# indicates that the candidate surrogate was generated using a censored normal distribution.$^\dagger$ indicates estimation under minor violation of assumption A6, where it is for the slightly modified interpretation of TPF and FPF (see section \ref{CAM})}
\end{tabular}
\end{center}
\end{sidewaystable}

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power to detect difference in $\widetilde{STG}$ \label{IPow}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}^{\#}=0.47$& $\widetilde{STG}^{\#*}=0.23$  & $\widetilde{STG}^{\#*}=0.52$\\
  \hline
$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 15.5 & 98.7 & 14.6 & 98.2 \\ 
&  0.5  & 5.7 & 13.7 & 93.4 & 9.7 & 88.3 \\ 
 $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 74.2 & 5.3 & 69.5 \\ 
&  0.5 &  & 5.5 & 64.5 & 5.6 & 49.1 \\ 
$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 86.1 & 14.1 \\ 
 & 0.5 &  &  & 5.2 & 80.7 & 7.5 \\ 
$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 5.1 & 79.1 \\ 
&  (0.5, 0.5) &  &  &  & 5.8 & 78.2 \\ 
$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.4 \\ 
 & (0.5, 0.5) &  &  &  &  & 5.6 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$. \# indicates that the candidate surrogate was generated using a censored normal distribution.}
\end{tabular}
\end{center}
\end{sidewaystable}

@@

The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001. The P-value for testing H0$_1$ for EGFR is 0.508. The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.06. The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of $<$ 0.001. Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone. The estimated $\widehat{\widetilde{STG}}$ is 0.37 for HBA1C alone and 0.12 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.46. Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.
